Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-08-26 paper/main.tex

The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
% \sa{Maybe cite Dotter 2011 for previous distance? }
% APW: It's just a qualitative statement to give a rough sense of its distance, so I'm not really quoting a measurement
Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
The detailed but still limited kinematic data for Pal 5 has made it a canonical example of tidal destruction and a useful tool for constraining Galactic structure.
For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
% \sa{if you say the "shape and mass", you can cite Pearson 15 too ;)}
% APW: I say "measure" not "rule out" :) - but I'll reassess when I do the shortening and see if it fits
Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.

@@

Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 10 RRLs in the cluster itself (3 type~\typeab\ and 7 type~\typec) and 15 in the cluster tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.27$, the typical ratio found Oosterhoff I populations \citep{Smith1995}.  
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.53~d$.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

@@

The Pal~5 cluster has 5 previous known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
% The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
% The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.
Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}.
More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016} to identify \todo{XX} candidate members of the Pal 5 cluster and tails based on sky position and distance alone.
However, without a published catalog, we cannot compare our sample to theirs.

@@

Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag. In PS1 it is classified as an \rrc~with a period of 0.3832276153~d. The period reported in \citet{Vivas2004} is an alias of this one, more likely to be the true period given the variability amplitude\footnote{\citet{Vivas2004} also cautions that derived periods for QUEST \rrc~stars are known to be prone to aliasing due to the observing cadence of $\sim1$~d.}. Therefore, we keep PS1's classification henceforth for RRL 422.

@@

Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646 and a similarly low amplitude $A_G=0.2$ and $A_g=0.3$ in both surveys. This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d. We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.

\begin{table*}[t]
\caption{RRL members of the Pal 5 stream.}\label{t:rrl_members}
\begin{footnotesize}
\begin{tabular}{llllllll}
\toprule
\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & sep  \\
\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
\midrule
4418920808077110784 &  V5 & 0.337934& 0.46 & RRc  & 20.73 & 0.9982 &   1.9 \\
4418920846732620032 &     & 0.643239& 0.27 & RRab & 21.71 & 0.9963 &   2.0 \\
4418914863842345856 &  V1 & 0.293229& 0.56 & RRc  & 20.36 & 0.9696 &   2.0 \\
4418726027016125056 &  V3 & 0.329948& 0.51 & RRc  & 20.65 & 0.9978 &   3.8 \\
4418725889577171328 &  V4 & 0.286366& 0.59 & RRc  & 20.28 & 0.9915 &   4.5 \\
4418731829516302848 &     & 0.28164 & 0.3  & RRc  & 18.14 & 0.9297 &   4.8 \\
4418913218870688768 &  V2 & 0.332473& 0.4  & RRc  & 20.33 & 0.9977 &   5.0 \\
4418734165978521728 &     & 0.380859& 0.52 & RRc  & 22.75 & 0.9850 &   7.6 \\
4418724034151291776 & 403 & 0.551705& 1.14 & RRab & 21.23 & 0.9987 &   9.7 \\
4418732791593809152 & 405 & 0.329670& 0.45 & RRc  & 20.74 & 0.9959 &  11.0 \\
4419052204012341760 &     & 0.530006& 1.01 & RRab & 20.84 & 0.9960 &  44.5 \\
4418142117622280192 & 393 & 0.462671& 1.36 & RRab & 19.95 & 0.9060 & 118.7 \\
6339498589346000768 &     & 0.320110& 0.46 & RRc  & 18.89 & 0.9934 & 217.7 \\
6339499379619987200 &     & 0.329577& 0.47 & RRc  & 19.35 & 0.9966 & 218.6 \\
6339478312804685824 &     & 0.465883& 1.12 & RRab & 19.91 & 0.9970 & 242.8 \\
4421078432143578496 &     & 0.520749& 1.11 & RRab & 21.98 & 0.9981 & 246.1 \\
6339398155830632448 &     & 0.688316& 0.75 & RRab & 19.82 & 0.9982 & 258.9 \\
4427253907921168000 &     & 0.496334& 1.35 & RRab & 21.81 & 0.9905 & 362.5 \\
4427220338456828416 &     & 0.600796& 0.81 & RRab & 21.98 & 0.9974 & 384.4 \\
4427234700827469952 &     & 0.517911& 1.24 & RRab & 21.94 & 0.9979 & 392.3 \\
6337302795905934336 &     & 0.519462& 1.02 & RRab & 17.29 & 0.9808 & 404.5 \\
6337233350579111680 &     & 0.542383& 0.96 & RRab & 18.98 & 0.9962 & 450.6 \\
4427646704153765632 &     & 0.385552& 0.37 & RRc  & 19.90 & 0.9429 & 456.5 \\
4424705647988505600 &     & 0.335502& 0.41 & RRc  & 20.87 & 0.9832 & 513.0 \\
4426221707021159424 &     & 0.533313& 1.14 & RRab & 21.97 & 0.9728 & 550.1 \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table*}

>>>>

The Pal 5 stream---at a heliocentric distance of $d \sim 20~\kpc$---was discovered using early SDSS imaging, which reached well below the main sequence turnoff of the stream over a large area surrounding the cluster \citep{Odenkirchen:2001, Rockosi:2002}.
% \sa{Maybe cite Dotter 2011 for previous distance? }
% APW: It's just a qualitative statement to give a rough sense of its distance, so I'm not really quoting a measurement
Since then, radial velocities have been measured for a handful of giant branch stars in the cluster and along the stream \citep{Odenkirchen:2002, Odenkirchen:2009, Ibata:2017}, and deeper imaging data has been used to map the tails at higher signal-to-noise \citep{Bernard:2016, Ibata:2016, Bonaca:2019}.
The detailed but still limited kinematic data for Pal 5 has made it a canonical example of tidal destruction and a useful tool for constraining Galactic structure.
For example, using the sky track and radial velocity information, the stream has been used to measure the enclosed mass of the Milky Way \citep{Kuepper:2015, Bovy:2016}.
Using deep photometry, the density variations along the stream have been used to place limits on the abundance of massive substructures in the Galactic halo \citep{Erkal:2017}.

@@

Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find 9 RRLs in the cluster itself (2 type~\typeab\ and 7 type~\typec) and 15 in the cluster's tails (11 type~\typeab\ and 4 type~\typec), all  with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.2$, a typical ratio found Oosterhoff I populations \citep{Smith1995}.  
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.32$ \citep{TODO}.
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

@@

The Pal~5 cluster has 5 previously known RRL stars (named V1--V5), all of type which are \typec, as reported in the \citet{Clement:2001} compilation of variable stars in Galactic globular clusters and dating back to \citet{SawyerHogg:1973}.
In their analysis of the halo density profile with RRLs, \citet{Vivas:2006} report an overdensity (``Group~6'') that they claim could be associated with Pal~5's tidal tails.
Two of these stars (IDs 403 and 405) are separated by just $\sim10'$ and $11'$ from the cluster's center, which places them within the cluster's Jacobi radius, and an additional star (star 393) outside of the Jacobi radius but at a similar distance could be associated with the stream \citep{Vivas:2006}. We confirm the three stars are highly probable members. More recently, \citet{Ibata:2017} use a prior iteration of the PS1 RRL catalog \citep{Hernitschek:2016}, which contained \emph{plausible} RRLs, to identify 6 and 12 candidate members of the Pal 5 cluster and tails, respectively, based on sky position and distance alone. However, without a published catalog, we cannot compare our sample to theirs.

% The QUEST IDs of the stars in their Group~6 are: 403, 405, 407, 411, 416 and 422.
% The first two stars --403 and 405-- are located at distances of  $9\farcm8$ and $11'$ from the cluster's center, which \citet{Vivas2004} notes places them within the cluster's tidal radius, assuming the value of $r_t\sim16'$ reported by \citet{Odenkirchen:2002}.

@@

%Star 422 is reported as an \rrc~in \citet{Vivas2004}, with a period of 0.61925~d and $V$ amplitude 0.51~mag. In PS1 it is classified as an \rrc~with a period of 0.3832276153~d. The period reported in \citet{Vivas2004} is an alias of this one, more likely to be the true period given the variability amplitude\footnote{\citet{Vivas2004} also cautions that derived periods for QUEST \rrc~stars are known to be prone to aliasing due to the observing cadence of $\sim1$~d.}. Therefore, we keep PS1's classification henceforth for RRL 422.

@@

Star \verb+source_id=4418731829516302848+ is reported in PS1 and \Gaia~as an \rrab~with a period around 0.646~d and similarly low amplitudes $A_G=0.2$ and $A_g=0.3$ in both surveys. This star is reported in CRTS \citep{Drake2014} (ID CSS\_J151623.2-000831) as an \rrc~with a period 0.28164~d. We adopt the CRTS period for this star in what follows, as it is based on many more observations ($\gtrsim200$) than \Gaia~and PS1's (about a dozen per filter in each case) and the \rrc~classification and short period are more likely for such a short amplitude star.

\begin{table*}[t]
\caption{RRL members of the Pal 5 stream.}\label{t:rrl_members}
\begin{footnotesize}
\begin{tabular}{llllllll}
\toprule
\Gaia~DR2               & Other & Period & Amp-$g$ & Type & D & member & separation  \\
\verb+source_id+        & ID   & (d)    & (mag)   &     &(kpc)& prob & (") \\
\midrule
4418920808077110784 & V5 & 0.337934 & 0.46 & RRc & 20.7 & 1.00 & 2.0 \\
4418920846732620032 &    & 0.643239 & 0.27 & RRab & 21.7 & 1.00 & 2.0 \\
4418914863842345856 & V1 & 0.293229 & 0.56 & RRc & 20.4 & 0.97 & 2.0 \\
4418726027016125056 & V3 & 0.329948 & 0.51 & RRc & 20.7 & 1.00 & 3.9 \\
4418725889577171328 & V4 & 0.286366 & 0.59 & RRc & 20.3 & 0.99 & 4.5 \\
4418913218870688768 & V2 & 0.332473 & 0.40 & RRc & 20.3 & 1.00 & 5.1 \\
4418734165978521728 &    & 0.380859 & 0.52 & RRc & 22.8 & 0.92 & 7.7 \\
4418724034151291776 &403 & 0.551705 & 1.14 & RRab & 21.2 & 1.00 & 9.8 \\
4418732791593809152 &405 & 0.329670 & 0.45 & RRc & 20.7 & 1.00 & 11.0 \\
4419052204012341760 &    & 0.530006 & 1.01 & RRab & 20.8 & 1.00 & 44.5 \\
4418142117622280192 &393 & 0.462671 & 1.36 & RRab & 20.0 & 0.92 & 118.8 \\
6339498589346000768 &    & 0.320110 & 0.46 & RRc & 18.9 & 0.99 & 217.8 \\
6339499379619987200 &    & 0.329577 & 0.47 & RRc & 19.4 & 1.00 & 218.6 \\
6339478312804685824 &    & 0.465883 & 1.12 & RRab & 19.9 & 1.00 & 242.9 \\
4421078432143578496 &    & 0.520749 & 1.11 & RRab & 22.0 & 1.00 & 246.1 \\
6339398155830632448 &    & 0.688316 & 0.75 & RRab & 19.8 & 1.00 & 258.9 \\
4427253907921168000 &    & 0.496334 & 1.35 & RRab & 21.8 & 0.99 & 362.6 \\
4427220338456828416 &    & 0.600796 & 0.81 & RRab & 22.0 & 1.00 & 384.4 \\
4427234700827469952 &    & 0.517911 & 1.24 & RRab & 21.9 & 1.00 & 392.4 \\
6337302795905934336 &    & 0.519462 & 1.02 & RRab & 17.3 & 0.89 & 404.6 \\
6337233350579111680 &    & 0.542383 & 0.96 & RRab & 19.0 & 0.93 & 450.7 \\
4427646704153765632 &    & 0.385552 & 0.37 & RRc & 19.9 & 0.81 & 456.6 \\
4424705647988505600 &    & 0.335502 & 0.41 & RRc & 20.9 & 0.99 & 513.0 \\
4426221707021159424 &    & 0.533313 & 1.14 & RRab & 22.0 & 0.93 & 550.1 \\
\bottomrule
\end{tabular}
\end{footnotesize}
\end{table*}
